---
title: "Supplementary material"
subtitle: "Language structure is influenced by the proportion of non-native speakers: A reply to Koplenig (2019)"
author: "Henri Kauhanen, Sarah Einhaus & George Walkden"
date: "15 December 2022"
output: pdf_document
numbersections: true
toc: true
papersize: a4
---

# Introduction

This supplementary materials document provides all the statistical analyses reported in the paper, plus additional analyses that include interaction effects and a second imputation analysis in which a single imputation model is used to regress both types of complexity.

Tested with R version 4.0.4.


# Prerequisites

The following R packages are required:

```{r, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(mice)
library(broom.mixed)
library(lattice)
library(effects)
library(gridExtra)
```


# Data preprocessing

The raw data resides in two files (Koplenig's original dataset and our additions) in the `../data` directory. We first merge these two datasets and carry out a few transformations that will facilitate data analysis.

```{r}
kop <- read.csv("../data/rsos181274supp2.csv", stringsAsFactors=FALSE)
new <- read.csv("../data/koplenig-reply.csv", stringsAsFactors=FALSE)
new <- new[, c("ISO", "ethnologue_L2_users", "used_as_L2_by", "notes")]
```

Format the data slightly differently:

```{r}
for (i in 1:nrow(new)) {
  tmp <- paste(stringr::str_extract_all(new[i, ]$used_as_L2_by,
                                        pattern="\\[[a-z]+\\]")[[1]], collapse=":")
  new[i, ]$used_as_L2_by <- stringr::str_replace_all(tmp, pattern="[\\[\\]]", "")
}
new$used_as_L2_by <- ifelse(new$used_as_L2_by=="", NA, new$used_as_L2_by)
```

Merge the two dataframes:

```{r}
data <- merge(kop, new, by="ISO")
```

For some languages, the area is missing, but these are read in as empty strings rather than as missing values. Need to fix that:

```{r}
data$Area <- ifelse(data$Area == "", NA, data$Area)
data$Family <- ifelse(data$Family == "", NA, data$Family)
```

Make sure language family and area are factors (important for imputation model and regression analysis):

```{r}
data$Family <- factor(data$Family)
data$Area <- factor(data$Area)
```

Encode the logarithm of population size and the logarithm of the range size as variables in the dataframe (useful for some of the regressions and plots):

```{r}
data$logPop <- log(data$Population)
data$logRangesize <- log(data$Rangesize)
```


# Descriptive statistics

## General characteristics of the dataset

There are a total of

```{r}
nrow(data)
```

languages in the dataset. However, not every language has data for each column of the data frame. The number of vehicular languages is

```{r}
nrow(data[data$vehicularity==1, ])
```

Of these,

```{r}
nrow(data[data$vehicularity==1 & is.na(data$L2prop), ])
```

do not have an L2 proportion estimate (either real or imputed).

The number of non-vehicular languages is
```{r}
nrow(data[data$vehicularity==0, ])
```

These all have an L2 proportion estimate, either real or imputed:

```{r}
nrow(data[data$vehicularity==0 & is.na(data$L2prop), ])
```



## How many non-vehiculars have an imputed L2 proportion?

The number of non-vehicular languages with a zero L2 proportion is

```{r}
nv0 <- nrow(data[data$vehicularity==0 & data$L2prop==0, ])
nv0
```

Of these, Ethnologue actually provides a numerical zero L2 proportion estimate for

```{r}
nv0E <- nrow(data[data$vehicularity==0 & data$L2prop==0 & 
             data$ethnologue_L2_users==TRUE, ])
nv0E
```

languages. The rest have been imputed.

## In how many cases is the data imputation wrong?

Ethnologue notes that the language is used as an L2 by speakers of some other set of languages (without giving numerical estimates) in

```{r}
asL2 <- nrow(data[data$vehicularity==0 & data$L2prop==0 & 
             !is.na(data$used_as_L2_by), ])
asL2
```

of these cases. In other words, the data imputation is definitely wrong for

```{r}
asL2/(nv0 - nv0E)
```

of the dataset.



# Remove zero-imputation from uncertain non-vehiculars

We now remove the zero-imputed L2 proportions from uncertain non-vehicular languages:

```{r}
data2 <- data
data2$L2prop <- ifelse(data2$vehicularity==0 & data2$L2prop==0 & 
                       data2$ethnologue_L2_users==FALSE, NA, data2$L2prop)
```

There are now

```{r}
nrow(data2[!is.na(data2$L2prop), ])
```

languages with a non-`NA` L2 proportion. Of these,

```{r}
nrow(data2[!is.na(data2$L2prop) & data2$vehicularity==1, ])
```

are vehicular and

```{r}
nrow(data2[!is.na(data2$L2prop) & data2$vehicularity==0, ])
```

non-vehicular.

We point out that there are missing values also in the response variables, morphological complexity and information-theoretic complexity. In other words, the two complexity measures are available for different subsets of languages:

```{r}
nrow(data2[!is.na(data2$MC), ])
nrow(data2[!is.na(data2$H), ])
```

In particular, in the subset of languages with a non-missing L2 proportion, these numbers are:

```{r}
nrow(data2[!is.na(data2$L2prop) & !is.na(data2$MC), ])
nrow(data2[!is.na(data2$L2prop) & !is.na(data2$H), ])
```


# Overall missingness in the data

The variables in the dataset now have this many missing values:

```{r}
nrow(data2[is.na(data2$Family), ])
nrow(data2[is.na(data2$Area), ])
nrow(data2[is.na(data2$MC), ])
nrow(data2[is.na(data2$H), ])
nrow(data2[is.na(data2$L2prop), ])
nrow(data2[is.na(data2$Population), ])
nrow(data2[is.na(data2$Rangesize), ])
```


# Family and area coverage

In the entire dataset, there are

```{r}
length(unique(data2$Family))
```

unique language families and

```{r}
length(unique(data2$Area))
```

unique linguistic areas. The three most frequent families have a fraction of

```{r}
sum(sort(as.numeric(table(data2$Family)), 
	 decreasing=TRUE)[1:3])/sum(!is.na(data2$Family))
```

of the languages. The three most frequent areas have a fraction of

```{r}
sum(sort(as.numeric(table(data2$Area)), 
	 decreasing=TRUE)[1:3])/sum(!is.na(data2$Area))
```

of the languages (not counting languages for which area is missing).

The above statistics for our reduced sample are:

```{r}
data2b <- data2
data2b <- data2b[!is.na(data2b$L2prop), ]
length(unique(data2b$Family))
length(unique(data2b$Area))
sum(sort(as.numeric(table(data2b$Family)), 
	 decreasing=TRUE)[1:3])/sum(!is.na(data2b$Family))
sum(sort(as.numeric(table(data2b$Area)), 
	 decreasing=TRUE)[1:3])/sum(!is.na(data2b$Area))
```


# Complete cases analysis

In the complete cases analysis, we do not impute any missing values.

## Morphological complexity

```{r}
mod <- lmer(MC~L2prop+logPop+(1|Family)+(1|Area), data2)
summary(mod)
```

Adding an interaction between L2 proportion and population size leads to a worse model:

```{r}
modb <- lmer(MC~L2prop*logPop+(1|Family)+(1|Area), data2)
AIC(mod)
AIC(modb)
```


## Morphological complexity, $\geq$ 6 features

When only looking at languages in which at least 6 features are available for the determination of morphological complexity, we cannot include the same random effects structure because it leads to a singular fit:

```{r}
mod6 <- lmer(MC~L2prop+logPop+(1|Family)+(1|Area), data2[data2$NumChap>=6, ])
```

Apparently, this is because the area is missing for many languages. Hence we run the following, simpler model instead:

```{r}
mod6 <- lmer(MC~L2prop+logPop+(1|Family), data2[data2$NumChap>=6, ])
summary(mod6)
```

Adding an interaction between L2 proportion and population size again leads to a worse model:

```{r}
mod6b <- lmer(MC~L2prop*logPop+(1|Family), data2[data2$NumChap>=6, ])
AIC(mod6)
AIC(mod6b)
```


## Information-theoretic complexity

For information-theoretic complexity, the random effects structure with a random intercept for area again leads to a singular fit:

```{r}
modIC <- lmer(H~L2prop+logPop+(1|Family)+(1|Area), data2)
```

Hence we only include a random intercept for family:

```{r}
modIC <- lmer(H~L2prop+logPop+(1|Family), data2)
summary(modIC)
```

Adding an interaction between L2 proportion and population size leads to a worse model:

```{r}
modICb <- lmer(H~L2prop*logPop+(1|Family), data2)
AIC(modIC)
AIC(modICb)
```


# Multiple imputation analysis

## Preparatory steps

We provide two different kinds of analysis: first, one which has separate imputation models for the two kinds of complexities, and second, an analysis which has a single imputation model for both complexities.

The rationale for constructing separate imputation models (the analysis reported in the main paper) is that the number of languages for which both morphological complexity and information-theoretic complexity are attested is rather small:

```{r}
nrow(data2[!is.na(data2$MC), ])
nrow(data2[!is.na(data2$H), ])
nrow(data2[!is.na(data2$MC) & !is.na(data2$H), ])
```

All of our imputation models take language family as a clustering variable. The implementation in *mice* requires this as a numeric:

```{r}
datai <- data2
datai$cluster <- as.numeric(datai$Family)
```

Also, we want to make sure that impossible L2 speaker proportions (outside the interval [0,1]) are never imputed. To do this, we take a logit transform of L2 speaker proportion:

```{r}
epsilon <- 10^-5
datai$L2prop_t <- epsilon + (1 - 2*epsilon)*datai$L2prop
datai$L2prop_t <- log(datai$L2prop_t/(1 - datai$L2prop_t))
```


## Separate imputation models

### Morphological complexity

We need the following variables in this imputation model:

```{r}
datai_MC <- datai[, c("ISO", "Language", "Family", "Area", "L2prop_t", "MC",
                      "logPop", "logRangesize", "cluster")]
```

We first set up the predictor matrix. L2 speaker proportion is imputed using morphological complexity, logarithmic population size and logarithmic range size, with language family as a clustering variable. We do not impute other missing values; we have tried to do so, but the model becomes too complicated to run.

```{r}
pred_MC <- make.predictorMatrix(datai_MC)
pred_MC[1:nrow(pred_MC), ] <- 0
pred_MC["L2prop_t", ] <- c(0, 0, 0, 0, 0, 1, 1, 1, -2)
pred_MC
```

We also need to set the imputation method:

```{r}
impmethod_MC <- character(ncol(datai_MC))
names(impmethod_MC) <- colnames(datai_MC)
impmethod_MC["L2prop_t"] <- "2l.lmer"
impmethod_MC
```

We now construct the imputation model. Note that we do not need more than one iteration, as missing values are only imputed in one variable.

```{r}
imp_MC <- mice(datai_MC, method=impmethod_MC, predictorMatrix=pred_MC, maxit=1, 
               m=100, print=FALSE, seed=202212)
```

Finally, we run the regression analysis on the $m = 100$ completed copies of the dataset:

```{r}
modImp <- with(imp_MC, lmer(MC~L2prop_t+logPop+(1|Family)+(1|Area)))
tidy(pool(modImp))
```

For purposes of illustration, here is the regression on one of the 100 completed datasets (that is, before pooling):

```{r}
summary(modImp$analyses[[1]])
```


### Information-theoretic complexity

We need the following variables in this imputation model:

```{r}
datai_H <- datai[, c("ISO", "Language", "Family", "Area", "L2prop_t", "H",
                     "logPop", "logRangesize", "cluster")]
```

We first set up the predictor matrix. L2 speaker proportion is imputed using information-theoretic complexity, logarithmic population size and logarithmic range size, with language family as a clustering variable. We do not impute other missing values; we have tried to do so, but the model becomes too complicated to run.

```{r}
pred_H <- make.predictorMatrix(datai_H)
pred_H[1:nrow(pred_H), ] <- 0
pred_H["L2prop_t", ] <- c(0, 0, 0, 0, 0, 1, 1, 1, -2)
pred_H
```

We also need to set the imputation method:

```{r}
impmethod_H <- character(ncol(datai_H))
names(impmethod_H) <- colnames(datai_H)
impmethod_H["L2prop_t"] <- "2l.lmer"
impmethod_H
```

We now construct the imputation model. Note that we do not need more than one iteration, as missing values are only imputed in one variable.

```{r}
imp_H <- mice(datai_H, method=impmethod_H, predictorMatrix=pred_H, maxit=1, 
              m=100, print=FALSE, seed=2202212)
```

Regression for information-theoretic complexity:

```{r}
modImpIC <- with(imp_H, lmer(H~L2prop_t+logPop+(1|Family)+(1|Area)))
tidy(pool(modImpIC))
```

For purposes of illustration, here is the regression on one of the 100 completed datasets (that is, before pooling):

```{r}
summary(modImpIC$analyses[[1]])
```


## Single imputation model

For the analysis with a single imputation model, we include both morphological and information-theoretic complexity:

```{r}
datai_s <- datai[, c("ISO", "Language", "Family", "Area", "L2prop_t", "MC",
                      "H", "logPop", "logRangesize", "cluster")]
```

We first set up the predictor matrix. L2 speaker proportion is imputed using morphological complexity, information-theoretic complexity, logarithmic population size and logarithmic range size, with language family as a clustering variable. We do not impute other missing values; we have tried to do so, but the model becomes too complicated to run.

```{r}
pred_s <- make.predictorMatrix(datai_s)
pred_s[1:nrow(pred_s), ] <- 0
pred_s["L2prop_t", ] <- c(0, 0, 0, 0, 0, 1, 1, 1, 1, -2)
pred_s
```

We also need to set the imputation method:

```{r}
impmethod_s <- character(ncol(datai_s))
names(impmethod_s) <- colnames(datai_s)
impmethod_s["L2prop_t"] <- "2l.lmer"
impmethod_s
```

We now construct the imputation model. Note that we do not need more than one iteration, as missing values are only imputed in one variable.

```{r}
imp_s <- mice(datai_s, method=impmethod_s, predictorMatrix=pred_s, maxit=1, 
              m=100, print=FALSE, seed=3202212)
```

Finally, we run the regression analyses on the $m = 100$ completed copies of the dataset:

```{r}
modImp_s <- with(imp_s, lmer(MC~L2prop_t+logPop+(1|Family)+(1|Area)))
tidy(pool(modImp_s))
modImpIC_s <- with(imp_s, lmer(H~L2prop_t+logPop+(1|Family)+(1|Area)))
tidy(pool(modImpIC_s))
```

Example regressions:

```{r}
summary(modImp_s$analyses[[1]])
summary(modImpIC_s$analyses[[1]])
```


# Plots

## Histogram of L2 speaker proportion

```{r, message=FALSE, warning=FALSE}
mypar1 <- list(par.main.text=list(just="left", x=grid::unit(21.5, "mm")))
mypar2 <- list(par.main.text=list(just="left", x=grid::unit(27.5, "mm")))
mycol <- "azure2"

g1 <- histogram(~L2prop, data2[!is.na(data2$L2prop) & data2$vehicularity==0, ], 
                type="percent", 
                col=mycol, xlab="L2 speaker proportion", par.settings=mypar1,
                main=list("Non-vehicular languages", cex=1.0), nint=13)

g2 <- histogram(~L2prop, data2[!is.na(data2$L2prop) & data2$vehicularity==1, ], 
                type="percent", 
                col=mycol, xlab="L2 speaker proportion", par.settings=mypar2,
                main=list("Vehicular languages", cex=1.0), nint=13)

# save as pdf
pdf("../plots/histogram.pdf", height=3, width=7)
grid.arrange(g1, g2, nrow=1, ncol=2)
dev.off()
```


## Effects plots (complete cases analysis)

```{r, message=FALSE, warning=FALSE}
g1 <- plot(predictorEffect("L2prop", mod), xlab="L2 speaker proportion", 
           ylim=c(0,1), ylab="Morphological complexity", 
           main="A                                  ")
g2 <- plot(predictorEffect("logPop", mod), xlab="log(population size)", 
           ylim=c(0,1), ylab="Morphological complexity", 
           main="B                                  ")
g3 <- plot(predictorEffect("L2prop", modIC), xlab="L2 speaker proportion", 
           ylim=c(1.3, 2.0), ylab="Information-theoretic complexity", 
           main="C                                  ")
g4 <- plot(predictorEffect("logPop", modIC), xlab="log(population size)", 
           ylim=c(1.3, 2.0), ylab="Information-theoretic complexity", 
           main="D                                  ")

# pdf out
pdf("../plots/result.pdf", height=7.5, width=7)
grid.arrange(g1, g2, g3, g4, nrow=2, ncol=2)
dev.off()
```


## Histogram of L2 proportion coefficients in imputation analysis

```{r, message=FALSE, warning=FALSE}
mypar1 <- list(par.main.text=list(just="left", x=grid::unit(21.5, "mm")))
mypar2 <- list(par.main.text=list(just="left", x=grid::unit(14.5, "mm")))
mycol <- "azure2"

getcoef <- function(X) {
  estimate <- coef(X)$Family$L2prop_t[1]
  estimate
}

d1 <- do.call(rbind, lapply(X=modImp$analyses, FUN=getcoef))
d2 <- do.call(rbind, lapply(X=modImpIC$analyses, FUN=getcoef))

g1 <- histogram(d1, type="percent", 
                xlab=expression("Coefficient estimate for"~rho*"'"), 
                col=mycol, par.settings=mypar1,
                main=list("Morphological complexity", cex=1.0), nint=13)

g2 <- histogram(d2, type="percent", 
                xlab=expression("Coefficient estimate for"~rho*"'"), 
                col=mycol, par.settings=mypar2,
                main=list("Information-theoretic complexity", cex=1.0), nint=13)

# save as pdf
pdf("../plots/imputation.pdf", height=3, width=7)
grid.arrange(g1, g2, nrow=1, ncol=2)
dev.off()
```


